---
title: "Why Data Cleaning Matters: Methods and Tools in R"
author: Isabel Correia von der Beeck (400431670)
date: '2025-11-26'
format: revealjs
html:
  embed-resources: true
header: "Why Data Cleaning Matters: Methods and Tools in R"
transition: slide
bibliography: mylit.bib
csl: _extensions/apaquarto/apa.csl
slide-number: true
---

```{r}
library(tidyverse)
library(gt)
library(readxl)
theme_set(theme_minimal())
Mappe1 <- read_excel("Mappe1.xlsx")
```

## Agenda

-   What is it good for?

-   Overview of methods & tools

-   Exercise



## What is Data Cleaning Good for?

**Key benefits:** 

-   **Accuracy** of results by removing errors, duplicates or other inconsistencies

-   **Consistency** across datasets makes it possible to merge and compare datasets without conflict

-   Handling **missing values** to prevent biased outcomes

-   Better modeling and **predictions** when using statistival models

*see* @kabacoff2024modern, @Lee2025guide 


## Overview 

| Package | Function  | Use                                         |
|---------|-----------|---------------------------------------------|
| dplyr   | select    | select variables/columns                    |
| dplyr   | filter    | select observations/rows                    |
| dplyr   | mutate    | transform or recode variables               |
| dplyr   | summarize | summarize data                              |
| dplyr   | group_by  | identify subgroups for further processing   |


*see* @nordmann2022data, @kabacoff2024modern

## dataset starwars

```{r| eval: false}
starwars %>%
  head(10)

```

```{r}
starwars %>%
  head(10)
```


## select( )
-   keeps only the columns you need, removes irrelevant ones
```{r| eval: false}
starwars %>%
select(name,species,contains("color"))  %>%
head(10)
```

```{r}
starwars %>%
select(name,species,contains("color"))  %>%
head(10)
```


## filter( )
-   keeps only rows that meet certain conditions
```{r| eval: false}
starwars %>%
  select(name, height, species, eye_color) %>%
  filter(species == "Human" &
           height < 200 &
           eye_color %in% c("blue", "brown")) %>%
  head(9)
```
```{r}
starwars %>%
  select(name, height, species, eye_color) %>%
  filter(species == "Human" &
           height < 200 &
           eye_color %in% c("blue", "brown")) %>%
  head(9)
```


## mutate( )
-   creates new variables or modifies existing ones
```{r| eval: false}
starwars %>%
  select(name, height, mass, species) %>%
  mutate(height = height/100) %>%
  head(10)
```

```{r}
starwars %>%
  select(name, height, mass, species) %>%
  mutate(height = height/100) %>%
  head(10)
```


## mutate( )
-   creates new variables or modifies existing ones
```{r| eval: false}
starwars %>%
  select(name, height, mass, species) %>%
  mutate(height_2 = height/100) %>%
  head(10)
```

```{r}
starwars %>%
  select(name, height, mass, species) %>%
  mutate(height_2 = height/100) %>%
  head(10)
```


## mutate( ) and recode ( )
-   replace or rename categories for consistency
```{r| eval: false}
starwars %>%
  select(name, contains("color"), species) %>%
  mutate(species = recode(species,
                          "Droid" = "Robot")) %>%
  head(10)

```

```{r}
starwars %>%
  select(name, contains("color"), species) %>%
  mutate(species = recode(species,
                          "Droid" = "Robot")) %>%
  head(10)
```


## missing data

```{r| eval: false}
starwars %>%
  select(name, gender, hair_color, height) %>%
  head(10)
```

```{r}
starwars %>%
  select(name, gender, hair_color, height) %>%
  head(10)
```
## missing data - na.omit( )
-   drops all rows with missing values
```{r| eval: false}
starwars %>%
  select(name, gender, hair_color, height) %>%
  na.omit()
head(10)
```

```{r}
starwars %>%
  select(name, gender, hair_color, height) %>%
  na.omit()
head(10)
```


## missing data - filter( )
-   shows all rows with missing value

```{r| eval: false}
starwars %>%
  select(name, gender, hair_color, height) %>%
  filter(!complete.cases(.))
```

```{r}
starwars %>%
  select(name, gender, hair_color, height) %>%
  filter(!complete.cases(.))
```
## missing data - drop.na( )
-   remove rows only if specific columns have missing values
```{r| eval: false}
starwars %>%
  select(name, gender, hair_color, height) %>%
  filter(!complete.cases(.)) %>%
  drop_na(height)
head(10)
```
```{r}
starwars %>%
  select(name, gender, hair_color, height) %>%
  filter(!complete.cases(.)) %>%
  drop_na(height)
head(10)
```
## missing data - replace_na( )
-   fills missing values with defaults
```{r| eval: false}
starwars %>%
  select(name, gender, hair_color, height) %>%
  filter(!complete.cases(.)) %>%
  mutate(hair_color = replace_na(hair_color, "none"))
```

```{r}
starwars %>%
  select(name, gender, hair_color, height) %>%
  filter(!complete.cases(.)) %>%
  mutate(hair_color = replace_na(hair_color, "none"))
```

## group_by ( ) and summarise ( )
-   organise data into groups (grouped operations) and collapse groups into summary statistics (e.g., mean, count, sum)
```{r| eval: false}
starwars %>%
  group_by(species) %>%
  summarise(avg_mass = mean(mass, na.rm = TRUE))

```
```{r}
starwars %>%
  group_by(species) %>%
  summarise(avg_mass = mean(mass, na.rm = TRUE))
head(10)
```


## Learning my Doing
[***Task:***]{.underline}***What is the average female and male height (in m) and mass in the "starwars" data set?***
```{r#| eval: false}
# install.packages("tidyverse")
library(tidyverse)
view(starwars)
```

***Try to use the functions:***
-   select()
-   filter()
-   mutate()
-   drop_na()
-   group_by()
-   summarise()

## Possible Solution

```{r#| eval: false}
starwars %>%
  select(sex, height, mass) %>%
  filter(sex == "male" | sex == "female") %>%
  mutate(height = height/100) %>%
  drop_na() %>%
  group_by(sex) %>%
  summarise('Average height' = (mean(height)),
            'Average mass' = mean(mass))
```

```{r}
starwars %>%
  select(sex, height, mass) %>%
  filter(sex == "male" | sex == "female") %>%
  mutate(height = height/100) %>%
  drop_na() %>%
  group_by(sex) %>%
  summarise('Average height' = (mean(height)),
            'Average mass' = mean(mass))
```


::: {.center .middle}
# Thank you for listening!
:::



## References

::: {#refs}
:::
